---
title: "NYPD Shooting Incident Data Report"
author: "Justin B"
date: "2026-05-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Packages

```{r Load Packages}
library(tidyverse)
```


## Data Import
Data obtained from **https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD** 

```{r Read CSV}
nypd_csv_path <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_data <- read_csv(nypd_csv_path, skip_empty_rows = TRUE)
```
```{r Data Display}
head(nypd_data)
```
```{r Data Str}
summary(nypd_data)
```
```{r Unique Race Values}
unique(nypd_data$VIC_RACE)
```


## Clean Data
```{r Clean Data}
# Get the number of PERP on VIC crime
nypd_data_clean <- nypd_data %>%
  filter(!is.na(PERP_RACE)) %>%
  filter(!is.na(VIC_RACE)) %>%
  filter(!(PERP_RACE %in% c('(null)', 'UNKNOWN'))) %>%
  filter(!(VIC_RACE %in% c('(null)', 'UNKNOWN'))) %>%
  mutate(PERP_ON_VIC_RACES = paste0(PERP_RACE, '>', VIC_RACE))

nypd_data_clean
```

## Data Analysis

### Murders by Borough

```{r}
murder_by_borough <-  nypd_data_clean %>%
  group_by(BORO) %>%
  count(BORO, name = 'COUNT')

murder_by_borough %>%
  ggplot(aes(x = BORO, y = COUNT)) +
  geom_bar(stat = 'identity') +
  theme_bw()
```

### Race on Race Crime
```{r Race on Race}
race_on_race <- nypd_data_clean %>%
  count(PERP_ON_VIC_RACES, name = 'COUNT') %>%
  arrange(desc(COUNT)) %>%
  head(10)

race_on_race %>%
  ggplot(aes(x = reorder(PERP_ON_VIC_RACES, -COUNT), y = COUNT)) +
  geom_bar(stat = 'identity') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  xlab('Perp>Victim Race')
```
What is the proportion on same-race versus difference-race race-on-race crime?
```{r}
nypd_data_clean %>%
  count(PERP_ON_VIC_RACES, name = 'COUNT') %>%
  arrange(desc(COUNT)) %>%
  mutate(PERP_RACE = gsub(">.*.", "", PERP_ON_VIC_RACES)) %>%
  mutate(VICTIM_RACE = gsub(".*.>", "", PERP_ON_VIC_RACES)) %>%
  mutate(RACE_ON_RACE_TYPE = ifelse(PERP_RACE == VICTIM_RACE, "same-race", "different-race")) %>%
  group_by(RACE_ON_RACE_TYPE) %>%
  summarize(count = sum(COUNT)) %>%
  mutate(prop = count / sum(count)) %>%
  ggplot(aes(x = RACE_ON_RACE_TYPE, y = prop)) +
  geom_bar(stat='identity')
```


## Data Modelling
Is the increased prevelance of same-race crime consistent across all boroughs?

```{r}
df <- nypd_data_clean %>%
  count(PERP_ON_VIC_RACES, BORO, name = "COUNT") %>%
  mutate(
    PERP_RACE   = sub(">.*", "", PERP_ON_VIC_RACES),
    VICTIM_RACE = sub(".*>", "", PERP_ON_VIC_RACES),
    RACE_ON_RACE_TYPE = ifelse(PERP_RACE == VICTIM_RACE,
                               "same-race", "different-race")
  )

head(df)
```

```{r}
model <- glm(
  COUNT ~ BORO + PERP_RACE + VICTIM_RACE + PERP_RACE:VICTIM_RACE,
  family = poisson,
  data = df
)

summary(model)
```
```{r, fig.width=10, fig.height=10}
df$expected <- predict(model, type = "response")
df$pearson_resid <- residuals(model, type = "pearson")

ggplot(df, aes(PERP_RACE, VICTIM_RACE, fill = pearson_resid)) +
  geom_tile() +
  facet_wrap(~ BORO) +
  theme_minimal() +
  coord_equal() +
  theme(axis.text.x.bottom = element_text(angle=90, hjust=1))
```

```{r}
ggplot(df, aes(expected, COUNT)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  theme_minimal()
```

Looks fairly 1-to-1 linear, which makes sense since we're comparing against the data 
we have trained again. To test this appropriately most likely would need to do cross-validation.

# Brief comment on Bias

The analysis may be biased towards certain demographics which have a larger proportion of crimes and 
skew the definition of same-race crime. For example, certain ethnicity in New York may be less prevalent and therefore less likely to conduct same-race crimes, or more likely to conduct different-race crime. It's important to consider this when conducting analysis like this.